### 2018-03-16 submission/manuscript.Rmd
a new re-write of the abstract

**Background.** Many of the different steps used during 16S rRNA gene sequencing can introduce bias and by extension change the observed microbial community results. Although cycle number and high fidelity (HiFi) DNA polymerase are studied less often, they are still important sources of bias in this workflow. Here, we critically examine both cycle number and HiFi DNA polymerase for biases that may influence downstream bacterial community results.

**Methods.** DNA from fecal samples (n = 4) were extracted using a PowerMag DNA extraction kit with a 10 minute bead beating step and amplified at 15, 20, 25, 30, and 35 cycles using Accuprime, Kappa, Phusion, Platinum, or Q5 HiFi DNA polymerase. Mock communities (technical replicates n = 4) consisting of previously isolated whole genomes of 8 different bacteria were also amplified using the same approach. First, the number of OTUs (Operational Taxonomic Units) were examined for both fecal samples and mock communities. Next, Bray-Curtis index, the error rate, sequence error prevalence, and chimera prevalence were investigated. Finally, the chimera prevalence correlation with number of OTUs was assessed.

**Results.** At 35 cycles there were significant differences between HiFi DNA polymerase for fecal samples (P-value `r ifelse(max(filter(fs_anova_summary, cycle == "35x")[, "bh"]) < 0.0001, invisible("<"), invisible("check data"))` 0.0001). These HiFi dependent differences in the number of OTUs could be identified as early as 20 cycles in the mock communities (P-value = `r format(filter(mock_anova_summary, sub_sample_level == 1000, cycle == "20x")[, "bh"], digits = 1, nsmall = 1)`). Chimera prevalence varied by HiFi DNA polymerase and these differences were still observed after chimera removal using VSEARCH. Additionally, the chimera prevalence had a strong positive correlation with the number of OTUs and this association was not changed by chimera removal with VSEARCH.

**Conclusions.** Due to HiFi DNA polymerase dependent differences in the number of OTUs and chimera prevalence, common diversity metrics could have values that are not comparable across studies.

>>>>

**Background.** Using different reagents and kits at different steps of the 16S rRNA gene sequencing workflow can introduce bias by changing the observed microbial community. Although cycle number and high fidelity (HiFi) DNA polymerase are studied less often, they are still important sources of bias in this workflow. Here, we examine how both cycle number and HiFi DNA polymerase  can change the bacterial community and introduce bias to the final obtained results.

**Methods.** DNA from fecal samples (n=4) were extracted using a PowerMag DNA extraction kit with a 10 minute bead beating step and amplified at 15, 20, 25, 30, and 35 cycles using Accuprime, Kappa, Phusion, Platinum, or Q5 HiFi DNA polymerase. Mock communities (technical replicates n=4) consisting of previously isolated whole genomes of 8 different bacteria were also amplified using the same approach. The number of OTUs (Operational Taxonomic Units) were examined for both fecal samples and mock communities. Next, Bray-Curtis index, the error rate, sequence error prevalence, and chimera prevalence were investigated. Finally, the chimera prevalence correlation with number of OTUs was assessed.

**Results.** When analyzing fecal samples a different number of OTUs was observed between HiFi DNA polymerases at 35 cycles (P-value `r ifelse(max(filter(fs_anova_summary, cycle == "35x")[, "bh"]) < 0.0001, invisible("<"), invisible("check data"))` 0.0001). These HiFi dependent differences in the number of OTUs were identified as early as 20 cycles in the mock communities (P-value = `r format(filter(mock_anova_summary, sub_sample_level == 1000, cycle == "20x")[, "bh"], digits = 1, nsmall = 1)`). Chimera prevalence varied by HiFi DNA polymerase and this variation was still present after chimera removal using VSEARCH. The chimera prevalence was positively correlated with the number of OTUs and was also not affected by chimera removal with VSEARCH.

**Conclusions.** HiFi DNA polymerase dependent differences in the number of OTUs and chimera prevalence makes comparison across studies difficult. Care should be exercised when choosing both HiFi DNA polymerase and cycle number to be used in 16S rRNA gene sequencing studies.

### 2018-03-16 submission/manuscript.tex
a new re-write of the abstract

\textbf{Background.} Many of the different steps used during 16S rRNA
gene sequencing can introduce bias and by extension change the observed
microbial community results. Although cycle number and high fidelity
(HiFi) DNA polymerase are studied less often, they are still important
sources of bias in this workflow. Here, we critically examine both cycle
number and HiFi DNA polymerase for biases that may influence downstream
bacterial community results.

\textbf{Methods.} DNA from fecal samples (n = 4) were extracted using a
PowerMag DNA extraction kit with a 10 minute bead beating step and
amplified at 15, 20, 25, 30, and 35 cycles using Accuprime, Kappa,
Phusion, Platinum, or Q5 HiFi DNA polymerase. Mock communities
(technical replicates n = 4) consisting of previously isolated whole
genomes of 8 different bacteria were also amplified using the same
approach. First, the number of OTUs (Operational Taxonomic Units) were
examined for both fecal samples and mock communities. Next, Bray-Curtis
index, the error rate, sequence error prevalence, and chimera prevalence
were investigated. Finally, the chimera prevalence correlation with
number of OTUs was assessed.

\textbf{Results.} At 35 cycles there were significant differences
between HiFi DNA polymerase for fecal samples (P-value \textless{}
0.0001). These HiFi dependent differences in the number of OTUs could be
identified as early as 20 cycles in the mock communities (P-value =
0.002). Chimera prevalence varied by HiFi DNA polymerase and these
differences were still observed after chimera removal using VSEARCH.
Additionally, the chimera prevalence had a strong positive correlation
with the number of OTUs and this association was not changed by chimera
removal with VSEARCH.

\textbf{Conclusions.} Due to HiFi DNA polymerase dependent differences
in the number of OTUs and chimera prevalence, common diversity metrics
could have values that are not comparable across studies.

>>>>

\textbf{Background.} Using different reagents and kits at different
steps of the 16S rRNA gene sequencing workflow can introduce bias by
changing the observed microbial community. Although cycle number and
high fidelity (HiFi) DNA polymerase are studied less often, they are
still important sources of bias in this workflow. Here, we examine how
both cycle number and HiFi DNA polymerase can change the bacterial
community and introduce bias to the final obtained results.

\textbf{Methods.} DNA from fecal samples (n=4) were extracted using a
PowerMag DNA extraction kit with a 10 minute bead beating step and
amplified at 15, 20, 25, 30, and 35 cycles using Accuprime, Kappa,
Phusion, Platinum, or Q5 HiFi DNA polymerase. Mock communities
(technical replicates n=4) consisting of previously isolated whole
genomes of 8 different bacteria were also amplified using the same
approach. The number of OTUs (Operational Taxonomic Units) were examined
for both fecal samples and mock communities. Next, Bray-Curtis index,
the error rate, sequence error prevalence, and chimera prevalence were
investigated. Finally, the chimera prevalence correlation with number of
OTUs was assessed.

\textbf{Results.} When analyzing fecal samples a different number of
OTUs was observed between HiFi DNA polymerases at 35 cycles (P-value
\textless{} 0.0001). These HiFi dependent differences in the number of
OTUs were identified as early as 20 cycles in the mock communities
(P-value = 0.002). Chimera prevalence varied by HiFi DNA polymerase and
this variation was still present after chimera removal using VSEARCH.
The chimera prevalence was positively correlated with the number of OTUs
and was also not affected by chimera removal with VSEARCH.

\textbf{Conclusions.} HiFi DNA polymerase dependent differences in the
number of OTUs and chimera prevalence makes comparison across studies
difficult. Care should be exercised when choosing both HiFi DNA
polymerase and cycle number to be used in 16S rRNA gene sequencing
studies.
